---
title: "chapter 2"
author: "Zachary Himmelberger"
output:
  html_document: default
  pdf_document: default
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```
Importing the rethinking package
```{R, message=FALSE}
library(rethinking)
```
##Chapter 2
#2E1
$$P(rain|Monday)$$
#2E2
The probability that it is Monday, given that it is raining.
#2E3
$$P(Monday|rain)$$
#2E4
In the globe tossing example, the amount of water on Earth is being estimated. The parameter of interest is a value that is exact. The uncertainty exists only in the model. Probability, therefore, is a function of our knowledge about the parameter.
#2M1
```{r}
p_grid <- seq(from=0, to=1, length.out=20)
prior <- rep(1,20)
likelihood <- dbinom(3, size=3, prob=p_grid)
unstd.posterior <- likelihood*prior
posterior <- unstd.posterior / sum(unstd.posterior)
plot(p_grid,posterior)
```
#2M2
```{R}
p_grid <- seq(from=0, to=1, length.out=20)
prior <- c(rep(0,10), rep(1,10))
likelihood <- dbinom(3, size=3, prob=p_grid)
unstd.posterior <- likelihood*prior
posterior <- unstd.posterior / sum(unstd.posterior)
plot(p_grid,posterior)
```
#2M3
$$P(E|land) = \frac{P(land|E) P(E)}{P(land|E) P(E) + P(land|M) P(M)}$$
```{R}
p.E.given.land <- (.3*.5) / (.3*.5 + 1*.5)
```
#2M4
W/W has 0 ways; W/B has 1 way; B/B has 2 ways
therefore, B/B has 2 of 3 ways or P(B/B) = .66
#2M5
W/W has 0 ways; W/B has 1 way; B/B has 2 ways; B/B has 2 ways
therefore, B/B has 4 of 5 ways or P(B/B) = .80
#2M6
W/W has 0 ways; W/B has 1 way; B/B has 2 ways
multiplied by prior probs results in 0*3; 1*2; 2*1
therefore, B/B has 2 of 4 ways or P(B/B) = .50

#2M7
probabilities of first card are
W/W has 0 ways; W/B has 2 ways; B/B has 6 ways
therefore, B/B has 6 of 8 ways or P(B/B) = .75

#2H1
$$P(sB|twins) = \frac{P(twins|sB) P(sB)}{P(twins|sB) P(sB) + P(twins|sA) P(sA)}$$
```{R}
p.sB.given.twins <- (.2*.5) / (.2*.5 + .1*.5)
```
$$P(twins) = P(twins|s1) P(s1) + P(twins|s2) P(s2)$$
```{R}
p.twins <- (.2*.66)+(.1*.33)
```
#2H2
$$P(sA|twins) = \frac{P(twins|sA) P(sA)}{P(twins|sA) P(sA) + P(twins|sB) P(sB)}$$
```{R}
p.sA.given.twins <- (.1*.5) / (.1*.5 + .2*.5)
```
#2H3
$$P(sA|single) = \frac{P(single|sA) P(sA)}{P(single|sA) P(sA) + P(single|sB) P(sB)}$$
```{R}
p.sA.given.single <- (.9*.33) / (.9*.33 + .8*.66)
```
#2H4
$$P(sA|testA) = \frac{P(testA|sA) P(sA)}{P(testA|sA) P(sA) + P(testA|sB) P(sB)}$$
```{R}
p.sA.given.testA <- (.8*.5) / (.8*.5 + .65*.5)
```
including new test information but using prior from above
```{R}
p.sA.given.testA.prior <- (.8*.36) / (.8*.36 + .65*.64)
```