---
title: "Chapter 2"
author: "Zachary Himmelberger"
date: "12/14/2021"
output: html_document
---

# Import Packages

```{r, message=FALSE}
library(rethinking)
```

# Practice Problems

## 2M1

### 1

```{r}
# define grid
p_grid <- seq(from=0, to=1, length.out=20)

# define prior
prior <- rep(1, 20)

# compute likelihood at each value in grid
likelihood <- dbinom(3, size = 3, prob = p_grid)

# compute product of likelihood and prior
unstd.posterior <- likelihood * prior

# standardize the posterior, so it sums to 1
posterior <- unstd.posterior / sum(unstd.posterior)

# plot the results
plot(p_grid, posterior, type = "b",
     xlab = "prob of water", ylab = "posterior prob")
```

### 2

```{r}
# define grid
p_grid <- seq(from=0, to=1, length.out=20)

# define prior
prior <- rep(1, 20)

# compute likelihood at each value in grid
likelihood <- dbinom(3, size = 4, prob = p_grid)

# compute product of likelihood and prior
unstd.posterior <- likelihood * prior

# standardize the posterior, so it sums to 1
posterior <- unstd.posterior / sum(unstd.posterior)

# plot the results
plot(p_grid, posterior, type = "b",
     xlab = "prob of water", ylab = "posterior prob")
```

### 1

```{r}
# define grid
p_grid <- seq(from=0, to=1, length.out=20)

# define prior
prior <- rep(1, 20)

# compute likelihood at each value in grid
likelihood <- dbinom(5, size = 7, prob = p_grid)

# compute product of likelihood and prior
unstd.posterior <- likelihood * prior

# standardize the posterior, so it sums to 1
posterior <- unstd.posterior / sum(unstd.posterior)

# plot the results
plot(p_grid, posterior, type = "b",
     xlab = "prob of water", ylab = "posterior prob")
```

## 2M2.

### 1

```{r}
# define grid
p_grid <- seq(from=0, to=1, length.out=20)

# define prior
prior <- ifelse(p_grid > .5, 1, 0)

# compute likelihood at each value in grid
likelihood <- dbinom(3, size = 3, prob = p_grid)

# compute product of likelihood and prior
unstd.posterior <- likelihood * prior

# standardize the posterior, so it sums to 1
posterior <- unstd.posterior / sum(unstd.posterior)

# plot the results
plot(p_grid, posterior, type = "b",
     xlab = "prob of water", ylab = "posterior prob")
```

## 2H1

First we calculate the probability of species A given twins and species B given twins.

\begin{align}
P(t | A) &= .1 \\
P(t | B) &= .2 \\
P(A | t) &= \frac{P(t | A) P(A)}{P(t)} \\
&= \frac{P(t | A) P(A)}{P(t | A) P(A) + P(t | B) P(B)} \\
&= \frac{.1 \times .5}{.1 \times .5 + .2 \times .5} \\
&= \frac{.05}{.15} \\
&= .33
\end{align}

Thus, $P(B | t) = .66$. Then, the probability of the next birth being twins is equal to

\begin{align}
P(t) &= P(A | t) P(A) + P(B | t) P(B) \\
&= .33 \times .1 + .66 \times .2 \\
&\approx .165
\end{align}

## 2H2

\begin{align}
P(t | A) &= .1 \\
P(t | B) &= .2 \\
P(A | t) &= \frac{P(t | A) P(A)}{P(t)} \\
&= \frac{P(t | A) P(A)}{P(t | A) P(A) + P(t | B) P(B)} \\
&= \frac{.1 \times .5}{.1 \times .5 + .2 \times .5} \\
&= \frac{.05}{.15} \\
&= .33
\end{align}

## 2H3

\begin{align}
P(A | t, s) &= \frac{P(t, s | A) P(A)}{P(t, s)} \\
&= \frac{.9 \times .33}{.9 \times .33 + .8 \times .66} \\
&\approx \frac{.33}{.858} \\
&\approx .385

\end{align}

## 2H4

First, we can determine the probability of species A given a positive test for species A and the probability of species B given a negative test for species B. Note that this makes additional assumptions about the test, but are supported by the answers given by McElreath.

\begin{align}
P(test A | A) &= .8 \\
P(test A | B) &= 1 - .65 \\
&= .35
\end{align}

\begin{align}
P(A | test A) &= \frac{P(test A | A) P(A)}{P(test A)} \\
&= \frac{.8 \times .5}{.8 \times .5 + .35 \times .5} \\
&= \frac{.4}{.575} \\
&= .70
\end{align}

Now plugging in the information about births, we can use an updated prior. 

\begin{align}
P(A | test A) &= \frac{P(test A | A) P(A)}{P(test A)} \\
&= \frac{.8 \times .36}{.8 \times .36 + .35 \times .64} \\
&\approx \frac{.288}{.512} \\
&\approx .563
\end{align}

Show a video and tell people it is a deep fake. Measure how it affects them.

Show prosthetic limb v. no limb and see if it affects
